### 2025-07-22 modern-portfolio-theory.qmd
20250721 - links

Modern portfolio theory (mean-variance theory) is a framework for determining the optimal composition of an investment portfolio to maximize expected returns for a given level of risk.
Here, *risk* refers to the [variability](#sec-descriptiveStatisticsSpread) (e.g., standard deviation or variance) of returns across time.
Given two portfolios with the same expected returns over time, people will prefer the "safer" portfolio—that is, the portfolio with less [variability](#sec-descriptiveStatisticsSpread)/volatility across time.
One of the powerful notions of modern portfolio theory is that, through diversification, one can achieve lower risk with the same expected returns.
In investing, *diversification* involves owning multiple asset classes (e.g., domestic and international stocks and bonds), with the goal of having asset classes that are either uncorrelated or negatively correlated.
That is, owning different types of assets is safer than owning only one type.
If you have too much money in one asset and that asset tanks, you will lose your money.
In other words, you do not want to put all of your eggs in one basket.
By owning different asset classes, you can limit your downside risk without sacrificing much in terms of expected return.

This lesson can also apply to fantasy football.
When assembling a team, you are essentially putting together a portfolio of assets (i.e., team of players).
As with stocks, each player has an expected return (i.e., projection) and a degree of risk.
In fantasy football, a player's risk might be quantified in terms of the [variability](#sec-descriptiveStatisticsSpread) of projected scores for a player across projection sources (e.g., Projection Source A, Source B, Souce C, etc.), or as historical game-to-game [variability](#sec-descriptiveStatisticsSpread).
Variability of projected scores for a player across projection sources could reflect the *uncertainty* of projections for a player.
Variability of historical (actual) fantasy points across games could reflect many factors, including risks due to injuries, situational changes (e.g., being traded to a new team or changes in team composition such as due to the acquisition of new players on the team), game scripts, and the tendency for the player to be "boom-or-bust" (e.g., if they are highly dependent on scoring touchdowns or long receptions for fantasy points).
All things equal, we want to minimize our risk for a given level of expected returns.
That way, we have the best chance of winning any given week.
For the same level of expected returns, higher risk teams might have a few amazing games, but their teams might fall flat in other weeks.
That is, for a given (high) rate of return, [you are best off in the long run (i.e., over the course of a season) with a lower risk team](https://eng.wealthfront.com/2012/01/17/moneyball-using-modern-portfolio-theory-to-win-your-fantasy-sports-league) compared to a higher risk team [@Hitchings2012; archived at <https://perma.cc/NE35-G6LR>].

@@

However, sometimes having two players on the same team might be beneficial because some positions may be uncorrelated or even negatively correlated, which can also reduce risk.
For instance, the performance of the Tight End and Running Back on the same team tends to be slightly negatively correlated, so it might not be a bad idea to start the Tight End and Running Back from the same team.
For a correlation matrix of all positions on the team, see: <https://assets-global.website-files.com/5f1af76ed86d6771ad48324b/607a4434a565aa7763bd1312_AndyAsh-Sharpstack-RPpaper.pdf> [@Sherman2021; archived at <https://perma.cc/JQ6G-KSRT>].

>>>>

Modern portfolio theory (mean-variance theory) is a framework for determining the optimal composition of an investment portfolio to maximize expected returns for a given level of risk.
Here, *risk* refers to the [variability](#sec-descriptiveStatisticsSpread) (e.g., standard deviation or variance) of returns across time.
Given two portfolios with the same expected returns over time, people tend to prefer the "safer" portfolio—that is, the portfolio with less [variability](#sec-descriptiveStatisticsSpread)/volatility across time.
One of the powerful notions of modern portfolio theory is that, through diversification, one can achieve lower risk with the same expected returns.
In investing, *diversification* involves owning multiple asset classes (e.g., domestic and international stocks and bonds), with the goal of having asset classes that are either uncorrelated or negatively correlated.
That is, owning different types of assets is safer than owning only one type.
If you have too much money in one asset and that asset tanks, you will lose your money.
In other words, you do not want to put all of your eggs in one basket.
By owning different asset classes, you can limit your downside risk without sacrificing much in terms of expected return.
In sum, the goal of diversification is to reduce risk (not to increase returns) and thus to increase risk-adjusted returns, by reducing the amount of risk needed for a given level of return.

This lesson can also apply to fantasy football.
When assembling a team, you are essentially putting together a portfolio of assets (i.e., team of players).
As with stocks, each player has an expected return (i.e., projection) and a degree of risk.
In fantasy football, a player's risk might be quantified in terms of the [variability](#sec-descriptiveStatisticsSpread) of projected scores for a player across projection sources (e.g., Projection Source A, Source B, Source C, etc.), or as historical game-to-game [variability](#sec-descriptiveStatisticsSpread).
Variability of projected scores for a player across projection sources could reflect the *uncertainty* of projections for a player.
Variability of historical (actual) fantasy points across games could reflect many factors, including risks due to injuries, situational changes (e.g., being traded to a new team or changes in team composition such as due to the acquisition of new players on the team), game scripts, and the tendency for the player to be "boom-or-bust" (e.g., if they are highly dependent on scoring touchdowns or long receptions for fantasy points).
All things equal, we want to minimize our risk for a given level of expected returns.
That way, we have the best chance of winning any given week.
For the same level of expected returns, higher risk teams might have a few amazing games, but their teams might fall flat in other weeks.
That is, for a given (high) rate of return, [you are best off in the long run (i.e., over the course of a season) with a lower risk team](https://eng.wealthfront.com/2012/01/17/moneyball-using-modern-portfolio-theory-to-win-your-fantasy-sports-league) compared to a higher risk team [@Hitchings2012; archived at <https://perma.cc/NE35-G6LR>].

@@

However, sometimes having two players on the same team might be beneficial because some positions may be uncorrelated or even negatively correlated, which can also reduce risk.
For instance, the performance of the Tight End and Running Back on the same team tends to be slightly negatively correlated, so it might not be a bad idea to start the Tight End and Running Back from the same team.
For a correlation matrix of all positions on the team, see: <https://assets-global.website-files.com/5f1af76ed86d6771ad48324b/607a4434a565aa7763bd1312_AndyAsh-Sharpstack-RPpaper.pdf> [@Sherman2021; archived at <https://perma.cc/JQ6G-KSRT>] and <https://www.4for4.com/2018/preseason/definitive-guide-stacking-draftkings> [@4for4Staff2018; archived at <https://perma.cc/JZH3-ZM5V>].
